---
title: "EDA Project Diamond Prices Part 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##1. Price vs. x

Create a scatterplot of Price vs X using the ggplot syntax

```{r, message = FALSE}
library(ggplot2)
```
```{r}
ggplot(aes(x = x, y = price), data = diamonds) +
  geom_point()
```

##2. Findings - Price vs. X

What are your observations about this scatterplot? **Exponential relationship with some outliers**

##3. Correlations

What is the corelation between price and x: **0.88**

What is the corelation between price and y: **0.87**

What is the corelation between price and z: **0.86**

```{r}
cor.test(diamonds$price, diamonds$x)
cor.test(diamonds$price, diamonds$y)
cor.test(diamonds$price, diamonds$z)
```

##4. Price vs Depth

Create a simple scatter plot of Price vs Depth

```{r}
ggplot(aes(x = depth, y = price), data = diamonds) +
  geom_point()
```

##5. Adjustments - Price vs. Depth

Change the code to make the transparency of the points to be 1/100 of what they are now and mark the x-axis every 2 units

```{r}
ggplot(data = diamonds, aes(x = depth, y = price)) + 
  geom_point(alpha = 1/100) +
  scale_x_continuous(breaks = seq(0,80,2))
```

##6. Typical Depth Range

Based on the Price vs Depth scatterplot, most diamonds are between what depths? **59 - 64**

##7. Correlation - Price and Depth

What is the correlation of depth vs price? **-0.01**

Would you use depth to predict price of a diamond? **No**

Why? The coefficient indicates there is no correlation between price and depth

```{r}
cor.test(diamonds$depth, diamonds$price)
```

##8. Price vs. Carat

Create a scatterplot of price vs carat and omit the top 1% of price and carat values

```{r}
ggplot(aes(x = carat, y = price), data = diamonds) + 
  geom_point() +
  xlim(0, quantile(diamonds$carat, 0.99)) +
  ylim(0, quantile(diamonds$price, 0.99))
```

##9. Price vs. Volume

Create a scatterplot of price vs. volume (x * y * z).This is a very rough approximation for a diamond's volume. 

Create a new variable for volume in the diamonds data frame.This will be useful in a later exercise.

```{r}
diamonds$volume <- diamonds$x * diamonds$y * diamonds$z
ggplot(aes(x = volume, y = price), data = diamonds) + 
  geom_point()
```

##10. Findings - Price vs. Volume

What are your observations of the price vs volume scatterplot? **very steep linear correlation with a few outliers, and a few null values for volume**

##11. Correlations on Subsets

What is the correlation of price and volume? Exclude diamonds with a volume of zero, or are greater than or equal to 800: **0.92**

```{r}
with(subset(diamonds, volume > 0 & volume < 800), cor.test(volume, price))
```

##12. Adjustments - Price vs. Volume

Subset the data to exclude diamonds with a volume greater than or equal to 800. Also, exclude diamonds with a volume of 0. Adjust the transparency of the points and add a linear model to the plot. 

Do you think this would be a useful model to estimate the price of diamonds? **Yes**

Why or why not? **High correlation between price and volume. However, other factors such as quality, color, clarity also play a role in price**

```{r}
diamonds_subset <- subset(diamonds, volume > 0 & volume < 800)
ggplot(aes(x = volume, y = price), data = diamonds_subset) + 
  ylim(0, 20000) +
  geom_point(alpha = 1/20) +
  geom_smooth(method = "lm", color = "blue")
```

##13. Mean Price by Clarity

Create a new data frame containing info on diamonds by clarity. Name the data frame diamondsByClarity. The data frame should contain the following variables in this order:

- (1) mean_price
- (2) median_price
- (3) min_price
- (4) max_price
- (5) n

Where n is the number of diamonds in each level of clarity.

```{r, message = FALSE}
library(dplyr)
```
```{r}
diamondsByClarity <- diamonds %>%
  group_by(clarity) %>%
  summarise(mean_price = mean(price),
            median_price = median(price),
            min_price = min(price),
            max_price = max(price),
            n = n()) %>%
  arrange(clarity)
head(diamondsByClarity)
```

##14. Bar Charts of Mean Price

Summary data frames with the mean price by clarity and color are provided.Write additional code to create two bar plots on one output image using the grid.arrange() function from the package gridExtra.

```{r}
diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, mean_price = mean(price))
diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, mean_price = mean(price))
```

```{r, message = FALSE}
library(gridExtra)
```
```{r}
grid.arrange(
ggplot(aes(x = diamonds_mp_by_clarity$clarity, y = diamonds_mp_by_clarity$mean_price), 
       data = diamonds_mp_by_clarity) +
  geom_bar(stat = "identity"),
ggplot(aes(x = diamonds_mp_by_color$color, y = diamonds_mp_by_color$mean_price), 
       data = diamonds_mp_by_color) +
  geom_bar(stat = "identity")
)
```

##15: Trends in Mean Price

What do you notice in each of the bar charts?

**The relationship between clarity and mean price, and color and mean price is not as expected - thought there would a stronger correlation between better clarity and color and higher price**

##16: Gapminder Revisited

Examine pairs of variables and create 2-5 plots that make use of the techniques from Lesson 4.

**Used economics data in ggplot as this lends itself to comparing two variables**

**Create a scatter plot the number of median duration of unemployment vs. unemployed population, with a mean line, and test the correlation.**
```{r}
ggplot(aes(x = unemploy, y = uempmed), data = economics) + 
  geom_point() +
  geom_smooth(method = "lm")
cor.test(economics$unemploy, economics$uempmed)
```
**The scatterplot indicates a high correlation between the unemployed population and the duration of unemployment, supported by a Pearson's coefficient of 0.87**

**Create a grouping of the data: for each personal savings rate, identify the mean personal consumption expenditure per population, the mean number of unemployed per population, and the count for each personal savings rate**
```{r}
economics.rate <- economics %>%
  group_by(psavert) %>%
  summarise(MeanPcePerPerson = mean(pce/pop),
            MeanUnemRate = mean(unemploy/pop),
            n = n())
head(economics.rate)
```

**Compare the means to the personal savings rates, and to each other**
```{r}
grid.arrange(
  ggplot(aes(x = MeanPcePerPerson, y = psavert), data = economics.rate) +
    geom_line() +
    geom_smooth(),
  ggplot(aes(x = MeanUnemRate, y = psavert), data = economics.rate) +
    geom_line() +
    geom_smooth(),
  ggplot(aes(x = MeanUnemRate, y = MeanPcePerPerson), data = economics.rate) +
    geom_line() +
    geom_smooth()
)
```

**The graphs indicate a strong negative correlation between personal savings rate and mean personal consumption expenditure per population, but weak correlations in the other two measures. This is confirmed by checking their correlation coefficients**

```{r}
cor.test(economics.rate$MeanPcePerPerson, economics.rate$psavert)
cor.test(economics.rate$MeanUnemRate, economics.rate$psavert)
cor.test(economics.rate$MeanUnemRate, economics.rate$MeanPcePerPerson)
```
